Temporal graph realization from fastest paths

### 2024-01-18 SAND2024/SAND24-longversion.tex
small changes to the long version

It is easy to see that the described algorithm runs in polynomial time. In the remainder, we proof that it is correct.

@@

Initially, our algorithm labels $e^\star$ with one, which equals $\lambda(e^\star)$. Now let $e$ be an edge popped off the queue by the algorithm in some iteration. Let $e'$ be the edge incident with $e$ that already obtained a label and is considered by the algorithm. Since $G$ is a tree, we have that $e'$ is closer to $e^\star$ than $e$. By induction we have that the algorithm labeled $e'$ with $\lambda(e')$. 
Assume that $e=\{u,v\}$ and $e'=\{v,w\}$. Since $G$ is a tree there is only one path from $u$ to $w$ in $G$ and it uses edges $e$ and $e'$. It follows that $\lambda(e')-\lambda(e)+1=D_{u,w}$ if $\lambda(e')>\lambda(e)$, and $\lambda(e')-\lambda(e)+\Delta+1=D_{u,w}$ otherwise. Our algorithm labels $e$ with $(\lambda(e')-D_{u,w}+1)\bmod \Delta$. It is straightforward to verify that the label of $e$ computed by the algorithm equals $\lambda(e)$. It follows that the algorithm computes $\lambda$.
\end{proof}

@@

Before presenting the structure of our algorithm for \cref{thm:FPTwrtFES}, observe that, in a static graph, the number of paths between two vertices can be upper-bounded by a function $f(k)$ of the feedback edge number $k$ of the graph. 
Therefore, for any fixed pair of vertices $u$ and $v$, we can ``guess'' the edges of the fastest temporal path from $u$ to $v$. 
However, 
% due to the difficulties illustrated in~\cref{fig:ftpExample}, 
for an FPT algorithm with respect to $k$, we cannot afford to guess the edges of the fastest temporal path for each of the $O(n^2)$ pairs of vertices. 
To overcome this difficulty, our algorithm follows this high-level strategy:
\begin{itemize}
    %%%\item We perform a preprocessing step by iteratively removing all vertices of degree one.
    \item We identify a small number $f(k)$ of ``important vertices''; these consist of the sets that we call $U, U^{\ast}, Z^{\ast}$.
    \item For each pair $u,v$ of important vertices, we guess the edges of the fastest temporal path from $u$ to $v$ (and from $v$ to $u$).
    \item From these guesses we can still not deduce the edges of the fastest temporal paths between many pairs of non-important vertices. However, as we prove, it suffices to guess only a small number of specific auxiliary structures (to be defined later).
    \item From these guesses we deduce fixed relationships between the labels of most of the edges of the graph. 
    \item For all the edges, for which we do not have deduced a label yet, we introduce a \emph{variable}. Using all these variables, we build an Integer Linear Program (ILP). 
    Among the constraints in this ILP we have that, for each of the $O(n^2)$ pairs of vertices $u,v$ in the graph, the duration of one specific temporal path from $u$ to $v$ (according to our guesses) is \emph{equal} to the desired duration $D_{u,v}$, while the duration of each of the other temporal paths from $u$ to $v$ is \emph{at least} $D_{u,v}$.
    \item 
    %%%By making any of the above guesses, we restrict the solution space for the problem \deltaExact. This restricted solution space coincides with the set of feasible solutions to the resulting ILP. Furthermore, the set of feasible solutions for all constructed ILPs coincide with the set of all solutions to \deltaExact\ (\ie regardless of our guesses). 
    By making each of the above combinations of guesses, we essentially enumerate all possible ways that our instance of \deltaExact\ has a solution, and for each of these possible ways we create an ILP. That is, our instance of \deltaExact\ has a solution if and only if at least one of these ILPs has a feasible solution.
    As each ILP can be solved in FPT time with respect to $k$ by Lenstra's Theorem~\cite{Lenstra1983Integer} (the number of variables is upper bounded by a function of $k$), we obtain our FPT algorithm for \deltaExact\ with respect to $k$.
\end{itemize}

@@

\subparagraph{\boldmath Fastest paths for vertices in $Z$.}
All of the above is enough to determine the labeling $\lambda$ of $G'$. Now we have to extend the labeling to consider also the vertices of~$Z= V(G) \setminus V(G')$ that we initially removed from $G$.
%
Now we are ready to explain how we determine the fastest temporal paths for all the vertices in $Z$

Recall that $G[Z]$ consists of disjoint trees and that each of these trees has a unique neighbor (clip vertex) $v$ in $G'$. We then define the tree $T_v$ in $G[Z \cup \{v\}]$ as the collection of trees from $G[Z]$ with a clip vertex $v$, together with the root $v$. 
Determining the fastest temporal paths between any two vertices in the same tree is a straightforward process (see~\ref{thm:deltaExact-PolyTimeTrees}), so we exclude this case from our upcoming analysis.
From \cref{obs:tree-only-neighbors} it follows that knowing temporal paths between any $y  \in V(G') \setminus (T_v)$ and all vertices in the first layer of the tree $T_v$ (\ie children of the root $v$),
it is enough to determine the fastest temporal paths between $y$ and all other vertices in $T_v$.
Therefore,
in the upcoming analysis, we focus only on the vertices in the first layer of each tree $T_v$.
During the process of determining the fastest paths from and to the vertices in $Z$, we use the fact that we have already identified the fastest paths among all vertices in $G'$.

@@

\subparagraph{\boldmath Fastest paths from $r \in Z$ to $y \in U \cup U^*$.}
Let $x \in V(G')$ be a clip vertex of the tree $T_x$ in $G[Z \cup \{x\}]$ with $r \in V(T_i)$ be a vertex in the first layer of $T_x$.
We distinguish the following two cases.
\begin{enumerate}[(i)]
    \item \label{FPT:clipInU-1}
    The clip vertex $x =u \in U$ is a vertex of interest.
    In this case we can w.l.o.g. assume that $r$ is a representative vertex in its equivalence class among the first layer vertices of $T_u$.
    From the guesses~\textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongUandZstar}}}} and \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongUstarandZstar}}}}
    we know the fastest temporal path from $r$ to $y$.
    \item \label{FPT:clipNotInU-1}
    The clip vertex $x \in U$ is not a vertex of interest. Then $x = v_j$ is a part of some segment $S_{u,v} = \{u=v_1, v_2, \dots, v_p = v)$, where $j \neq 1 \neq p$.
    Then all the labels of the tree $T_x$ are determined with respect to the label $\lambda(v_{j-1}v_j)$ and with respect to the label $\lambda(v_{j}v_{j+1})$ (see~\cref{lemma:treeLabels-notInteresting}).
    From the calculations above and the performed guesses we know the exact structure (\ie the sequence of vertices and edges) of the following paths:
    \begin{itemize}
        \item path $P_{xy}^*$ which is the fastest temporal path from the vertex $x$ to the vertex $y$,
        \item path $P_{xy}^u$ which is the fastest temporal path from the vertex $x$ to the vertex $y$, that passes through the vertex $u$,
        \item path $P_{xy}^v$ which is the fastest temporal path from the vertex $x$ to the vertex $y$, that passes through the vertex $v$.
    \end{itemize}
    Note that $P_{xy}^*$ is either equal to the path $P_{xy}^u$ or to the path $P_{xy}^v$.
    More precisely, from the guesses performed we know the structure of the fastest path from $v_2$ through $u$, that then continues to any other vertex of interest, and any other neighbor of the vertex of interest (see~guesses \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongv2z2}}}} and
    \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT:guess-uToSegmentz2}}}}).
    This path can then be easily (uniquely) extended to start from $x=v_i$, as there is a unique (temporal) path starting at $x$ and finishing in $u$ or $v$.

    Before we proceed note also that the labels of the tree are determined with respect to the label of the edge $v_{i-1}v_i$ and $v_{i}v_{i+1}$ (see~\cref{lemma:treeLabels-notInteresting}), therefore we can determine the difference 
    $|\lambda(v_{i-1} x) - \lambda (r x)|$ and $|\lambda(x v_{i+1}) - \lambda (r x)|$.
    %
    Suppose now that $P_{xy}^* = P_{xy}^u$.
    We calculate the value $c = D_{x,y} + |\lambda(v_{i-1} x) - \lambda (r x)|$ and compare it to $D_{r,y}$.
    There are three options.
    First $c = D_{r,y}$, in this case the fastest temporal path from $r$ to $y$ uses first the edge $rx$ and then continues to $y$ using the edges and vertices of $P_{xy}^*$.
    Second $c > D_{r,y}$, in this case the fastest temporal path from $r$ to $y$ uses first the edge $rx$ and then continues to $y$ using the vertices and edges of $P_{xy}^v$.
    Third $c < D_{r,y}$, in this case we stop the calculation and return false, as it cannot happen that a temporal path has a smaller duration than the corresponding value in the matrix $D$.
%
\end{enumerate}
In both cases, we introduce the equality constraint for the determined fastest temporal path and inequality constraint for all the other $k^{O(k)}$ paths.

\subparagraph{\boldmath Fastest paths from $r \in Z$ to $y \in V(G) \setminus (U \cup U^*)$.}
The proof in this case is similar to the one above.  We still split the analysis into two parts, one where the clip vertex $x$ of a tree $T_x$ that includes $r$ is in $U$ and one where it is not in $U$.
The difference is that in some cases we need to also extend the ending part of the path (which can be done uniquely, using the same arguments as in the previous case).

Once we determine the fastest temporal path from $r$ to $y$ we introduce an equality constraint for it, and for all other $k^{O(k)}$ paths we introduce the inequality constraints.

>>>>

It is easy to see that the described algorithm runs in polynomial time. In the remainder, we prove that it is correct.

@@

Initially, our algorithm labels $e^\star$ with one, which equals $\lambda(e^\star)$. Now let $e$ be an edge popped off the queue by the algorithm in some iteration, that is on the distance $i$ from $e^\star$. Let $e'$ be the edge incident with $e$ that is on the distance $i-1$ from $e^\star$. Since $G$ is a tree $e'$ has already been considered by the algorithm and thus already has a label. By induction we have that the algorithm labeled $e'$ with $\lambda(e')$. 
Assume that $e=\{u,v\}$ and $e'=\{v,w\}$. Since $G$ is a tree there is only one path from $u$ to $w$ in $G$ and it uses edges $e$ and $e'$. It follows that $\lambda(e')-\lambda(e)+1=D_{u,w}$ if $\lambda(e')>\lambda(e)$, and $\lambda(e')-\lambda(e)+\Delta+1=D_{u,w}$ otherwise. Our algorithm labels $e$ with $(\lambda(e')-D_{u,w}+1)\bmod \Delta$. It is straightforward to verify that the label of $e$ computed by the algorithm equals $\lambda(e)$. It follows that the algorithm computes $\lambda$.
\end{proof}

@@

Before presenting the structure of our algorithm for \cref{thm:FPTwrtFES}, observe that, in a static graph, the number of paths between two vertices can be upper-bounded by a function $f(k)$ of the feedback edge number $k$ of the graph.
This is true as any such path can traverse $0,1,2,\dots k$ feedback edges in different orders.
Therefore, for any fixed pair of vertices $u$ and $v$, we can ``guess'' the edges of the fastest temporal path from $u$ to $v$
(by guess we mean enumerate and test all possibilities).
However, 
% due to the difficulties illustrated in~\cref{fig:ftpExample}, 
for an FPT algorithm with respect to $k$, we cannot afford to guess the edges of the fastest temporal path for each of the $O(n^2)$ pairs of vertices. 
To overcome this difficulty, our algorithm follows this high-level strategy:
\begin{itemize}
    %%%\item We perform a preprocessing step by iteratively removing all vertices of degree one.
    \item We identify a small number $f(k)$ of ``important vertices''; these consist of the sets that we call $U, U^{\ast}, Z^{\ast}$.
    \item For each pair $u,v$ of important vertices, we guess the edges of the fastest temporal path from $u$ to $v$ (and from $v$ to $u$).
    \item From these guesses we can still not deduce the edges of the fastest temporal paths between many pairs of non-important vertices. However, as we prove, it suffices to guess only a small number of specific auxiliary structures (to be defined later).
    \item From these guesses we deduce fixed relationships between the labels of most of the edges of the graph. 
    \item For all the edges, for which we do not have deduced a label yet, we introduce a \emph{variable}. Using all these variables, we build an Integer Linear Program (ILP). 
    Among the constraints in this ILP we have that, for each of the $O(n^2)$ pairs of vertices $u,v$ in the graph, the duration of one specific temporal path from $u$ to $v$ (according to our guesses) is \emph{equal} to the desired duration $D_{u,v}$, while the duration of each of the other temporal paths from $u$ to $v$ is \emph{at least} $D_{u,v}$.
    \item 
    %%%By making any of the above guesses, we restrict the solution space for the problem \deltaExact. This restricted solution space coincides with the set of feasible solutions to the resulting ILP. Furthermore, the set of feasible solutions for all constructed ILPs coincide with the set of all solutions to \deltaExact\ (\ie regardless of our guesses). 
    By making each of the above combinations of guesses, we essentially enumerate all possible ways that our instance of \deltaExact\ has a solution, and for each of these possible ways we create an ILP. That is, our instance of \deltaExact\ has a solution if and only if at least one of these ILPs has a feasible solution.
    As each ILP can be solved in FPT time with respect to $k$ by Lenstra's Theorem~\cite{Lenstra1983Integer} (the number of variables is upper bounded by a function of $k$), we obtain our FPT algorithm for \deltaExact\ with respect to $k$.
\end{itemize}

@@

\subparagraph{\boldmath Fastest paths for vertices in $Z$.}
All of the above is enough to determine the labeling $\lambda$ of $G'$. We have to extend the labeling to consider also the vertices of~$Z= V(G) \setminus V(G')$ that we initially removed from $G$.

Recall that $G[Z]$ consists of disjoint trees and that each of these trees has a unique neighbor (clip vertex) $v$ in $G'$. We then define the tree $T_v$ in $G[Z \cup \{v\}]$ as the collection of trees from $G[Z]$ with a clip vertex $v$, together with the root $v$. 
Determining the fastest temporal paths between any two vertices in the same tree is a straightforward process (see~\cref{thm:deltaExact-PolyTimeTrees}), therefore we exclude this case from our upcoming analysis.
From \cref{obs:tree-only-neighbors} it follows that knowing temporal paths between any $y  \in V(G')$ and all vertices in the first layer of the tree $T_v$ (\ie children of the root $v$),
it is enough to determine the fastest temporal paths between $y$ and all other vertices in $T_v$.
Therefore,
in the upcoming analysis, we focus only on the vertices in the first layer of each tree $T_v$.
During the process of determining the fastest paths from and to the vertices in $Z$, we use the fact that we have already identified the fastest paths among all vertices in $G'$.

@@

\subparagraph{\boldmath Fastest paths from $r \in Z$ to $y \in U \cup U^*$.}
Let $x \in V(G')$ be a clip vertex of the tree $T_x$ in $G[Z \cup \{x\}]$ with $r \in V(T_i)$ be a vertex in the first layer of $T_x$.
We distinguish the following two cases.
\begin{enumerate}[(i)]
    \item \label{FPT:clipInU-1}
    The clip vertex $x =u \in U$ is a vertex of interest.
    In this case we can w.l.o.g. assume that $r$ is a representative vertex in its equivalence class among the first layer vertices of $T_u$.
    From the guesses~\textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongUandZstar}}}} and \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongUstarandZstar}}}}
    we know the fastest temporal path from $r$ to $y$.
    \item \label{FPT:clipNotInU-1}
    The clip vertex $x \in U$ is not a vertex of interest. Then $x = v_j$ is a part of some segment $S_{u,v} = \{u=v_1, v_2, \dots, v_p = v)$, where $j \neq 1 \neq p$.
    Using \cref{lemma:treeLabels-notInteresting} we can determine all the edge labels of $T_x$ with respect to the label $\lambda(v_{j-1}v_j)$ and with respect to the label $\lambda(v_{j}v_{j+1})$.
    Using the calculations of fastest temporal paths among vertices in $G'$ and the performed guesses we know the exact structure (\ie the sequence of vertices and edges) of the following paths:
    \begin{itemize}
        \item path $P_{xy}^*$ which is the fastest temporal path from the vertex $x$ to the vertex $y$,
        \item path $P_{xy}^u$ which is the fastest temporal path from the vertex $x$ to the vertex $y$, that passes through the vertex $u$,
        \item path $P_{xy}^v$ which is the fastest temporal path from the vertex $x$ to the vertex $y$, that passes through the vertex $v$.
    \end{itemize}
    Note that $P_{xy}^*$ is either equal to the path $P_{xy}^u$ or to the path $P_{xy}^v$.
    More precisely, from the guesses performed we know the structure of the fastest path from $v_2$ through $u$, that then continues to any other vertex of interest, and any other neighbor of the vertex of interest (see~guesses \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT-guessFTPamongv2z2}}}} and
    \textcolor{lipicsGray}{\textsf{\textbf{G-\ref{FPT:guess-uToSegmentz2}}}}).
    This path can then be easily (uniquely) extended to start from $x=v_i$, as there is a unique (temporal) path starting at $x$ and finishing in $u$ or $v$.

    Suppose now that $P_{xy}^* = P_{xy}^u$.
    Since the labels of $T_x$ are determined with respect to $\lambda(v_{i-1} x)$ we can calculate the value $c = D_{x,y} + |\lambda(v_{i-1} x) - \lambda (r x)|$.
    We then compare $c$ to $D_{r,y}$ and get one of the following three options.
    First $c = D_{r,y}$, in this case the fastest temporal path from $r$ to $y$ uses first the edge $rx$ and then continues to $y$ using the edges and vertices of $P_{xy}^*$.
    Second $c > D_{r,y}$, in this case the fastest temporal path from $r$ to $y$ uses first the edge $rx$ and then continues to $y$ using the vertices and edges of $P_{xy}^v$.
    Third $c < D_{r,y}$, in this case we stop the calculation and return false, as it cannot happen that a temporal path has a smaller duration than the corresponding value in the matrix $D$.
%
\end{enumerate}
In both cases, we introduce an equality constraint for the determined fastest temporal path and inequality constraints for all the other $k^{O(k)}$ paths.

\subparagraph{\boldmath Fastest paths from $r \in Z$ to $y \in V(G) \setminus (U \cup U^*)$.}
The proof in this case is similar to the one above.  We still split the analysis into two parts, one where the clip vertex $x$ of a tree $T_x$ that includes $r$ is in $U$ and one where it is not in $U$.
The difference is that in some cases we need to also extend the ending part of the path (which can be done uniquely, using the same arguments as in the above analysis).

Once we determine the fastest temporal path from $r$ to $y$ we introduce an equality constraint for it, and for all other $k^{O(k)}$ paths we introduce inequality constraints.

### 2024-01-18 SAND2024/SAND24-shortversion.tex
small changes to the long version

Initially, our algorithm labels $e^\star$ with one, which equals $\lambda(e^\star)$. Now let $e$ be an edge popped off the queue by the algorithm in some iteration. Let $e'$ be the edge incident with $e$ that already obtained a label and is considered by the algorithm. Since $G$ is a tree, we have that $e'$ is closer to $e^\star$ than $e$. By induction we have that the algorithm labeled $e'$ with $\lambda(e')$. 

>>>>

Initially, our algorithm labels $e^\star$ with one, which equals $\lambda(e^\star)$. Now let $e$ be an edge popped off the queue by the algorithm in some iteration, that is on the distance $i$ from $e^\star$. Let $e'$ be the edge incident with $e$ that is on the distance $i-1$ from $e^\star$. Since $G$ is a tree $e'$ has already been considered by the algorithm and thus already has a label. By induction we have that the algorithm labeled $e'$ with $\lambda(e')$. 
